---
title: "Decline on the Path to Recovery: Analysis"
author: "Li, Y-W, JW Malcom, JP Che-Castaldo, and MC Neel"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


<!--   rmarkdown::word_document: -->
<!--     toc: true -->
<!--     toc_depth: 3 -->
<!-- vignette: > -->
<!--   %\VignetteIndexEntry{Vignette Title} -->
<!--   %\VignetteEngine{knitr::rmarkdown} -->
<!--   %\VignetteEncoding{UTF-8} -->


## Introduction

This vignette covers the analysis of data on Endangered Species Act-listed 
species whose populations (number of populations or number of individuals) can
decline--relative to the population at time of listing and/or time of recovery
plan writing--and still be considered 'recovered.' A manuscript is in 
preparation.

```{r setup, include = FALSE}
library(dplyr)
library(decline.recovery)
library(ggplot2)
library(ggthemes)
library(stringr)

data("decline_recovery_final")
dat <- decline_recovery_final
rm(decline_recovery_final)

dat$Taxon <- gsub(dat$Taxon, pattern = "BIRD", replacement = "Bird")
dat$Taxon <- gsub(dat$Taxon, pattern = "FISH", replacement = "Fish")
dat$Taxon <- gsub(dat$Taxon, pattern = "MAMMAL", replacement = "Mammal")
dat$Taxon <- gsub(dat$Taxon, pattern = "REPTILE", replacement = "Reptile")

all_plans_years <- readRDS("../data-raw/recplan_by_year.rds")
names(all_plans_years) <- c("year", "n_all")
```

## Data filtering

We need to first filter out the species without a decline on the path to recovery. 
This includes one manual removal, Lake Erie water snake (_Nerodia sipedon 
insularum_), because a more refined read that suggests it doesn't qualify.

```{r filter}
filt <- filter(dat, !is.na(reason_lower_N_pop) |
                    !is.na(reason_lower_N_pop_sufficient) |
                    !is.na(reason_lower_N_ind) |
                    !is.na(reason_lower_N_ind_sufficient))
filt <- filter(filt, Scientific_Name != "Nerodia sipedon insularum")
dim(filt)

filt$pop_list2rec <- filt$N_pops_req_recovery / filt$N_pops_at_listing
filt$pop_plan2rec <- filt$N_pops_req_recovery / filt$N_pops_at_plan
filt$ind_list2rec <- filt$N_ind_req_recovery / filt$N_ind_at_listing
filt$ind_plan2rec <- filt$N_ind_req_recovery / filt$N_ind_at_plan

ind_lis <- filter(filt, filt$ind_list2rec < 1)
ind_pln <- filter(filt, filt$ind_plan2rec < 1)
pop_pln <- filter(filt, filt$pop_plan2rec < 1)
pop_lis <- filter(filt, filt$pop_list2rec < 1)
pop_lis_pln <- filter(filt, filt$pop_list2rec < 1 & filt$pop_plan2rec < 1)
ind_lis_pln <- filter(filt, filt$ind_list2rec < 1 & filt$ind_plan2rec < 1)
lis_pop_ind <- filter(filt, filt$pop_list2rec < 1 & filt$ind_list2rec < 1)
pln_pop_ind <- filter(filt, filt$ind_plan2rec < 1 & filt$pop_plan2rec < 1)
popind_lispln <- filter(filt, filt$ind_list2rec < 1 & 
                          filt$ind_plan2rec < 1 & 
                          filt$pop_list2rec < 1 & 
                          filt$pop_plan2rec < 1)

filt$pop_qual <- NA
for(i in 1:length(filt$pop_list2rec)) {
  if(is.na(filt$pop_list2rec[i]) & is.na(filt$pop_plan2rec[i])) {
    filt$pop_qual[i] <- NA
  } else if(is.na(filt$pop_list2rec[i]) & filt$pop_plan2rec[i] < 1) {
    filt$pop_qual[i] <- 1
  } else if(is.na(filt$pop_plan2rec[i]) & filt$pop_list2rec[i] < 1) {
    filt$pop_qual[i] <- 1
  } else if(is.na(filt$pop_list2rec[i]) & filt$pop_plan2rec[i] >= 1) {
    filt$pop_qual[i] <- 0
  } else if(is.na(filt$pop_plan2rec[i]) & filt$pop_list2rec[i] >= 1) {
    filt$pop_qual[i] <- 0
  } else if(filt$pop_plan2rec[i] < 1 | filt$pop_list2rec[i] < 1) {
    filt$pop_qual[i] <- 1
  } else {
    filt$pop_qual[i] <- 0
  }
}

filt$ind_qual <- NA
for(i in 1:length(filt$ind_list2rec)) {
  if(is.na(filt$ind_list2rec[i]) & is.na(filt$ind_plan2rec[i])) {
    filt$ind_qual[i] <- NA
  } else if(is.na(filt$ind_list2rec[i]) & filt$ind_plan2rec[i] < 1) {
    filt$ind_qual[i] <- 1
  } else if(is.na(filt$ind_plan2rec[i]) & filt$ind_list2rec[i] < 1) {
    filt$ind_qual[i] <- 1
  } else if(is.na(filt$ind_list2rec[i]) & filt$ind_plan2rec[i] >= 1) {
    filt$ind_qual[i] <- 0
  } else if(is.na(filt$ind_plan2rec[i]) & filt$ind_list2rec[i] >= 1) {
    filt$ind_qual[i] <- 0
  } else if(filt$ind_plan2rec[i] < 1 | filt$ind_list2rec[i] < 1) {
    filt$ind_qual[i] <- 1
  } else {
    filt$ind_qual[i] <- 0
  }
}

table(filt$pop_qual, useNA = "always")
table(filt$ind_qual, useNA = "always")

table(filt$pop_qual, filt$ind_qual,
      useNA = "always",
      dnn = c("Pop.", "Indiv."))

```

`r dim(dat)[1] - dim(filt)[1]` species dropped seems about right...which species?

```{r dropped}
drop <- setdiff(dat$Scientific_Name, filt$Scientific_Name)
drop
```

## Types of decline

```{r type_decline}
type_decl <- gsub(
  filt$Type_of_metric_with_decline,
  pattern = "#Ind",
  replacement = "# Ind"
)

type_decl <- gsub(
  type_decl,
  pattern = "#Pop",
  replacement = "# Pop"
)

type_decl <- gsub(
  type_decl,
  pattern = "writing,",
  replacement = "writing;"
)

type_decl <- gsub(
  type_decl,
  pattern = "listing vs delisting",
  replacement = "from listing"
)

type_decl <- gsub(
  type_decl,
  pattern = "Indv ",
  replacement = "Indivs "
)

type_decl <- gsub(
  type_decl,
  pattern = "Pops-",
  replacement = "Pops -"
)

types <- unlist(strsplit(type_decl, split = "; "))
table(types)

filt$Type_of_metric_with_decline <- type_decl
table(filt$Type_of_metric_with_decline)


new_types <- ifelse(
  grepl(type_decl, pattern = "Indivs") & grepl(type_decl, pattern = "Pop"), 
  "Ind & Pop",
  ifelse(grepl(type_decl, pattern = "Indivs"), "Ind", "Pop")
  ) 
table(new_types)

```

A question came up during the manuscript writing about which species have individuals decline
and an explanation for sufficiency (shown in Table 2 of the ms.). That can be found with:

```{r}
filter(ind_dec, reason_lower_N_ind == "Yes" & ind_qual == 1)$Scientific_Name
```

## Proportions for recovery

We can calculate the proportion of individuals / populations required at 
recovery, relative to the numbers at listing and plan writing.

```{r proportions}
filt$prop_ind_at_listing <- filt$N_ind_req_recovery / filt$N_ind_at_listing
filt$prop_ind_at_plan <- filt$N_ind_req_recovery / filt$N_ind_at_plan
filt$prop_pop_at_listing <- filt$N_pops_req_recovery / filt$N_pops_at_listing
filt$prop_pop_at_plan <- filt$N_pops_req_recovery / filt$N_pops_at_plan

summary(filt$prop_ind_at_listing)
summary(filt$prop_ind_at_plan)
summary(filt$prop_pop_at_listing)
summary(filt$prop_pop_at_plan)

props <- data_frame(
  proportions = c(
    filt$prop_ind_at_listing,
    filt$prop_ind_at_plan,
    filt$prop_pop_at_listing,
    filt$prop_pop_at_plan
  ),
  demography = c(
    rep("Individuals", 2 * length(filt$prop_ind_at_listing)),
    rep("Populations", 2 * length(filt$prop_ind_at_listing))
  ),
  timing = c(
    rep("Listing", length(filt$prop_ind_at_listing)),
    rep("Plan Writing", length(filt$prop_ind_at_listing)),
    rep("Listing", length(filt$prop_ind_at_listing)),
    rep("Plan Writing", length(filt$prop_ind_at_listing))
  )
)

ggplot(data = props, aes(proportions)) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = Inf),
            fill = "pink", alpha = 0.03) +
  geom_histogram() +
  scale_x_log10(breaks = c(0, 0.1, 1, 10, 100)) +
  labs(x = "Proportion (individuals or populations)", y = "Number of Species") +
  theme_hc() +
  facet_grid(timing ~ demography, scales = "free")
```

```{r taxo_dist}
tdf <- table(filt$Taxon) %>% 
  as.data.frame(stringsAsFactors = FALSE) %>%
  arrange(-Freq) %>%
  mutate(Var1=factor(Var1, levels=Var1))

ggplot(data = tdf, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "No. Species") +
  theme_hc()
```

Even though we have `r dim(filt)[1]` species that decline in some way, many of
these species also have substantial increases either in pops or individuals. 
But the bulk of proportional changes are small, so zooming in a bit:

```{r hist_zoom}
par(mfrow = c(2,2))
hist(filt$prop_ind_at_listing[filt$prop_ind_at_listing < 5], main = "")
hist(filt$prop_ind_at_plan[filt$prop_ind_at_plan < 5], main = "")
hist(filt$prop_pop_at_listing[filt$prop_pop_at_listing < 5], main = "")
hist(filt$prop_pop_at_plan[filt$prop_pop_at_plan < 5], main = "")
par(mfrow = c(1,1))
```

```{r prop_scatter}
qplot(data = filt, 
      x = prop_ind_at_listing, 
      y = prop_ind_at_plan, 
      geom = "point") +
  theme_bw()

qplot(data = filt, 
      x = prop_pop_at_listing, 
      y = prop_pop_at_plan, 
      geom = "point") +
  theme_bw()

qplot(data = filt, 
      x = prop_ind_at_listing, 
      y = prop_pop_at_listing, 
      geom = "point") +
  theme_bw()

qplot(data = filt, 
      x = prop_ind_at_plan, 
      y = prop_pop_at_plan, 
      geom = "point") +
  theme_bw()

```

The wide range of proportional values makes some of these graphs a little hard 
to parse. But one thing that is clear is that the proportions relative to
listing/plan writing are mostly independent.

## Taxonomic groups

Is there variation in the taxonomic groups represented, or how they qualify for
'decline on the path to recovery'?

```{r taxa}
table(filt$Taxon)
```

We could do a chi-square test, but the taxonomic distribution is clearly not
representative of listed species: plants (~50% of listed species) make up 
82% of species that can decline in abundance and still be recovered.

```{r taxa_2}
df <- data_frame(species = rep(filt$Scientific_Name, 4),
                 taxon = rep(filt$Taxon, 4),
                 metric = c(rep("ind_list", length(filt$Taxon)),
                            rep("ind_plan", length(filt$Taxon)),
                            rep("pop_list", length(filt$Taxon)),
                            rep("pop_plan", length(filt$Taxon))),
                 prop = c(filt$prop_ind_at_listing,
                          filt$prop_ind_at_plan,
                          filt$prop_pop_at_listing,
                          filt$prop_pop_at_plan))

aggregate(df$prop ~ df$taxon, FUN = median)
aggregate(df$prop ~ df$metric, FUN = median)
```

There's a lot of variation in median proportions between taxonomic groups and 
the metric of decline (or rather, change from listing / plan writing). More
interesting is the two-way breakdown:

```{r taxa_3}
combo_md <- aggregate(df$prop ~ df$metric + df$taxon, FUN = median)
combo_n <- aggregate(df$prop ~ df$metric + df$taxon, 
                     FUN = function(x) length(!is.na(x)))
names(combo_md) <- c("metric", "taxon", "prop")
names(combo_n) <- c("metric", "taxon", "N")
combo <- full_join(combo_md, combo_n)
combo
```

We can see that for most groups, the median proportion is < 1 for numbers of
populations rather than for numbers of individuals. In fact, most median 
proportions by taxon:metric are > 1, i.e., on 'average' the number of
individuals goes up even if the species qualifies based on a decline in the
number of populations. But we have to note that the sample size is pretty darn
small for most taxon:metric combinations, which really limits inference.

## Explanations of lower numbers

```{r reasons_1}
table(filt$reason_lower_N_pop, filt$reason_lower_N_pop_sufficient, useNA = "always")
table(filt$reason_lower_N_ind, filt$reason_lower_N_ind_sufficient, useNA = "always")
```

Still have more Explicit + Implicit > Yes (explained)...which ones?

```{r reason_mismatch}
mm_pop <- filter(filt, reason_lower_N_pop == "No" & 
                   reason_lower_N_pop_sufficient != "No")
mm_pop$Scientific_Name

mm_ind <- filter(filt, reason_lower_N_ind == "No" & 
                   reason_lower_N_ind_sufficient != "No")
mm_ind$Scientific_Name
```

~~~~It's unclear how the plan authors were able to suggest sufficiency of the 
numbers without acknowledging the decline.~~~~

How many plans had reasoning for both population- and individual-based numbers?

```{r combination_pop_ind}
twoway <- table(filt$reason_lower_N_pop, 
      filt$reason_lower_N_ind, 
      useNA = "always",
      dnn = c("Pop.", "Indiv."))
twoway
```

Only `r sum(twoway[1:2,1:2])` species qualified on both population- and individual-based numbers,
and no species included a reason for both types of decline.

How often was threat amelioration given as the reason the lower numbers were
sufficient for recovery?

```{r threat_amelior}
table(filt$threat_amelior_suffic)
```

Even though FWS often gives threat amelioration as the reason numbers can decline
and a species still be recovered, there's usually no mention of the idea and we
never found the argument made explicitly in these plans.

## Separate threats and demography criteria and actions?

Recovery plans would ideally provide separate delisting criteria for demographic
traits and for threats because doing so clarifies the reasons for listing and 
path to delisting. In addition, we would like to see recovery actions explicitly
described in terms of whether they are targeting demography or threats.

```{r threat_demog}
table(filt$threat_demog_explicit_criteria)
table(filt$threat_demog_explicit_actions)

table(filt$threat_demog_explicit_criteria, 
      filt$threat_demog_explicit_actions,
      dnn = c("criteria", "actions"))
```

## Three Rs

The three Rs -  resiliency, redundancy, and representation - describe a 
generally accepted principle of the requirements for recovery. It is particularly
important that the species with a decline on the path to recovery ensure that
each R is fulfilled. The first step in meeting the three Rs is discussing them
in the recovery plans.

```{r threeR}
table(filt$resiliency)
table(filt$redundancy)
table(filt$representation)

filt$any_3R <- ifelse(filt$resiliency == "No" &
                        filt$redundancy == "No" &
                        filt$representation == "No",
                      "No",
                      "Yes")
table(filt$any_3R)

table(filt$resiliency, 
      filt$redundancy, 
      useNA = "always",
      dnn = c("resil.", "redund."))
table(filt$resiliency, 
      filt$representation, 
      useNA = "always",
      dnn = c("resil.", "represent."))
table(filt$redundancy, 
      filt$representation, 
      useNA = "always",
      dnn = c("redund.", "represent."))
```

And now a three-way table, even though it is a little awkward:

```{r threeway_threeR}
table(filt$resiliency, 
      filt$redundancy, 
      filt$representation,
      dnn = c("resil.", "redund.", "represent."))
```

No species' plan discussed all three Rs explicitly, but the majority (`r 11+76`)
discuss each R at least implicitly and only seven species' plans do not address 
any of the Rs.

## Science of decline

That is, models used? (Note that some plans use a combination of model types.)

```{r models}
# First some additional text cleaning
mods <- filt$pop_models
mods <- str_replace_all(mods, "verbal", "Verbal")
mods <- str_replace_all(mods, "models|[Mm]odel", "")
mods <- str_replace_all(mods, "literature", "Literature")
mods <- str_replace_all(mods, "references|reference", "")
mods <- str_replace_all(mods, " $", "")
mod2 <- str_split(mods, " & ")
mod3 <- str_replace_all(unlist(mod2), " $", "")
table(unlist(mod3))
```

## Plan years

```{r plan_year}
qp <- qplot(filt$plan_year, geom = "histogram", bins = 32) + 
  labs(x = "Plan Year", y = "# Species") +
  theme_hc()
qp
# plotly::ggplotly(qp)

hist(filt$plan_year, breaks = 4, plot = FALSE)
# hist(filt$plan_year, breaks = 4)

dec_year <- table(filt$plan_year) %>% as.data.frame(stringsAsFactors = FALSE)
names(dec_year) <- c("year", "decline_n")
combo <- left_join(dec_year, all_plans_years, by = "year") %>%
  mutate(prop_decline = decline_n / n_all)
combo$year <- as.numeric(combo$year)

ggplot(combo, aes(x = year, y = prop_decline)) +
# qplot(data = combo, x = year, y = prop_decline, geom = "point") +
  geom_point(size = 4) +
  labs(x = "", y = "Proportion of Plans with Decline") +
  theme_hc()
  
```

### Changes through time

Now to break some measures down by block of time...

```{r thru_time}
filt$yr_block <- ifelse(
  filt$plan_year < 1990,
  1,
  ifelse(
    filt$plan_year < 2000,
    2,
    ifelse(
      filt$plan_year < 2010,
      3,
      4
    )
  )
)
```

The following get the numbers we need, but it's unclear how we illustrate
any of the interesting changes. Perhaps the good news (for me) is that there
are relatively few interesting results that might be considered 
interesting...most are tenuous at best because of the concentration of plans
from the 1990s.

#### No real change through time:

```{r}
tapply(filt$reason_lower_N_ind,
       INDEX = filt$yr_block,
       FUN = table)
```

#### Appears to shift towards 'yes', but tough to tell:

```{r}
tapply(filt$reason_lower_N_pop,
       INDEX = filt$yr_block,
       FUN = table)
```

#### No obvious shift:

```{r}
tapply(filt$reason_lower_N_ind_sufficient,
       INDEX = filt$yr_block,
       FUN = table)
```

#### Maybe a shift away from No:

```{r}
tapply(filt$reason_lower_N_pop_sufficient,
       INDEX = filt$yr_block,
       FUN = table)
```

#### Maybe a shift from No:

```{r}
tapply(filt$resiliency,
       INDEX = filt$yr_block,
       FUN = table)
```

#### Shift away from No:

```{r}
tapply(filt$redundancy,
       INDEX = filt$yr_block,
       FUN = table)
```


#### Less of a shift for representation:

```{r}
tapply(filt$representation,
       INDEX = filt$yr_block,
       FUN = table)
```

#### No strong shift:

```{r}
tapply(filt$threat_amelior_suffic,
       INDEX = filt$yr_block,
       FUN = table)
```

#### Looks like a shift away from combined threats & demography:

```{r}
tapply(filt$threat_demog_explicit_criteria,
       INDEX = filt$yr_block,
       FUN = table)
```

#### No change:

```{r}
tapply(filt$threat_demog_explicit_actions,
       INDEX = filt$yr_block,
       FUN = table)
```

## Conclusions

...



